### 2024-01-01 tesis_Ashish/Chapter7/chapter7.tex
Updated thesis 1 Jan 2024

The Phase II upgrade of the CMS Tracker is a crucial project undertaken to maintain the effectiveness of the CMS experiment during the High Luminosity LHC (HL-LHC) phase, which is expected to start in 2029. The HL-LHC will significantly increase the instantaneous luminosity, leading to a higher number of proton-proton collisions per bunch crossing. Instantaneous luminosity will increase to an unprecedented value of $7.5 \times 10^{34} cm^{-2} \: s^{-1}$ which corresponds to 200 proton-proton collisions per bunch crossing. This higher collision rate poses several challenges, including increased radiation levels, higher data rates, and more complex event environment. Run 2 pixel detector will not be able to handle these extreme radiation environment, resolve nearby particle tracks and operate properly to give a reliable estimate of the instantaneous luminosity for such a high pileup. That is why it will be replaced by a new pixel detector. The Phase II upgrade of the CMS tracker comprises of two main components: the Inner Tracker (IT) and the Outer Tracker (OT) \cite{RoyChowdhury:2729279}. The Inner Tracker consists of three main sub-components: the Tracker Endcap Pixel detector (TEPX), the Tracker Forward Pixel detector (TFPX), and the Tracker Barrel Pixel (TBPX) detector as shown in Fig. \ref{fig:Innertracker} \cite{CERN-LHCC-2017-009}.  %The Outer Tracker is responsible for detecting charged particles in the outer region of the CMS detector. It consists of two key subsystems: the silicon strip tracker and the silicon macro pixel tracker \cite{RoyChowdhury:2729279}.

Data Acquisition (DAQ) system for the Phase II CMS Tracker must handle the increased data rates and complexity associated with the high-luminosity phase of the LHC (HL-LHC). The HL-LHC is expected to achieve a peak luminosity which is around 5-7 times higher than the current LHC design luminosity which the DAQ system must handle. The upgraded CMS Tracker will generate a large amount of data due to the increased granularity of the pixel and strip sensors. The data rates from the tracker modules are expected to be several terabits per second (Tbps) during the HL-LHC operation. The Level-1 Trigger system in the PhaseII Upgrade is designed to accept events at a rate of up to 750 kHz. The DAQ system must handle the transmission of trigger primitives to the Level-1 Trigger system and buffer the full event data until the trigger decision is made. The high-speed optical links used for data transmission between the Front-End Electronics (FEE) and off-detector electronics are expected to operate at data rates of 10-25 Gbps or higher to handle the large data volumes generated by the upgraded tracker.

The TBPX occupies the central region along the \( z \)-axis from 0 to just below 300~mm, employing 360 units of \( 1 \times 2 \) modules (modules with two readout chips) and 504 units of \( 2 \times 2 \) modules (modules with four readout chips). The TFPX extends from approximately 300~mm to 600~mm in the \( z \)-direction, with 832 units of \( 1 \times 2 \) and 896 units of \( 2 \times 2 \) modules. The TEPX covers a broader range from about 600~mm to over 2500~mm and includes 768 units of \( 1 \times 2 \) and 992 units of \( 2 \times 2 \) modules.

The TEPX consists of four double disks per side with each double disk consisting of five rings of modules as shown in Fig. \ref{fig:Innertracker_40}. The rings have 20, 28, 36, 44 and 48 modules respectively.  The first two inner rings are equipped with \(1 \times 2\) modules, while the more extensive \(2 \times 2\) modules are mounted on the three outer rings. This design is further enhanced by the unique arrangement of modules in four layers as shown in Fig.   \ref{fig:Innertracker_41}. This allows for modules to be mounted on both the front and back sides of a disk, creating an overlap in the r-\(\phi\) direction. Additionally, by mounting modules at varying radii on the two disk that form each double-disk, overlap in the radial direction (r) is also accomplished.

@@

TEPX will have better radiation tolerance, increased granularity, improved two-track separation, improved estimation of hit rate and statistical precision, extending tracking acceptance beyond $|\eta|=4$. Collectively, these sub-detectors boast around 1.947 billion pixels across 13,488 pixel chips, spanning a silicon area of 4.87 square meters.

@@

\begin{figure}[!htp]
  \centering
  \includegraphics[width=1 \columnwidth]{ashish_thesis/tepx_tt_1.png}
  \caption[Phase II CMS Inner Tracker]{ \onehalfspacing A layout of the CMS Phase II Inner Tracker showing four TEPX disks, eight TFPX disks and four TBPX layers.}
  \label{fig:Innertracker}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=0.5 \columnwidth]{ashish_thesis/tepx_DD.png}
  \caption[TEPX double disk]{ \onehalfspacing View of one TEPX disk in the $r-\phi$ plane; individual pixel modules on two planes (double-disk) and five rings are outlined by yellow rectangles. The planes are separated in z by about 1 cm. The first plane holds modules on both front and back faces, with alternating rings to cover the gaps in r; the second plane holds modules in the same way as the first plane, but with an offset in $\phi$ to cover the gaps in $\phi$.}
  \label{fig:Innertracker_40}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=0.1 \columnwidth]{ashish_thesis/tepx_alllayer.png}
  \caption[TEPX double disk]{ \onehalfspacing R-Z view of the TEPX one double disk showing all four layers L1, L2, L3 and L4. Layer 1 and 3 has Ring 1, 3 and 5 modules, Layer 2 and 4 has Ring 2 and Ring 4 modules.}
  \label{fig:Innertracker_41}
\end{figure}

@@

TEPX is designed to function as a high-precision online bunch-by-bunch luminometer, primarily uses pixel cluster counting for luminosity measurement, a method involving the detection and counting of pixel clusters formed by charged particles from the proton collisions. Additionally, TEPX benefits from coincidence counting due to overlapping modules which enhances accuracy and enables systematic studies. %Positioned as part of the forward extension of the Inner Tracker detector,
TEPX features a sophisticated dedicated trigger system for sampling all relevant bunch crossings in an unbiased manner. This system operates at a baseline trigger rate of 75 kHz, which is 10$\%$  of the 750 kHz allocated for physics data. %In terms of data acquisition,
TEPX adapts to different operational conditions: during standard physics runs, it maintains a baseline luminosity trigger rate of 75 kHz, while during van der Meer (vdM) scans, this rate can significantly increase, potentially up to 1 MHz. TEPX readout rate (E-link speed) is estimated to be approximately 1330 Gbps at a pileup of 200  \cite{Collaboration:2706512}.
%The DAQ system of TEPX is designed to handle substantial data rates, with link bandwidth requirements varying from about 79.2 Gb/s at a trigger rate of 825 kHz during nominal physics operations to up to 24 Gb/s at the standard rate of 75 kHz  \cite{Collaboration:2706512}.

TEPX Disk 4 Ring 1 (D4R1) is located in the innermost ring of the last double-disk of the TEPX, is uniquely designed for measuring luminosity and beam-induced background (BIB). D4R1 is optimally placed due to its low radius, large active area, and high z position, making it highly sensitive and efficient for these measurements. The D4R1 operational model is distinct from the rest of TEPX \cite{Haranko2023}. It employs the entire available trigger rate and readout bandwidth, which varies based on operational mode (binary readout or Time-over-Threshold information). This capability ranges from a trigger rate of 825 kHz (similar to the rest of TEPX) up to several MHz, limited by readout bandwidth and the available e-link capacity. In nominal physics conditions, the baseline trigger rate for D4R1 is 825 kHz, while in van der Meer (vdM) conditions, it is 2 MHz, with the potential to reach between 5 and 10 MHz in certain scenarios.

@@

 Simulated data samples for Phase II include full CMS detector description and uses official CMS software (version $\texttt{CMSSW\_10\_6\_0\_patch2}$) which calls GEANT4 for detector simulation \cite{GEANT4:2002zbu}. Pixel cluster reconstruction is done by converting raw data to digi format (row, column, charge) and then form clusters from digis. The samples used in simulation contain single-neutrino event overlaid with a variable number of minimum-bias events (events with any amount of real energy detected in CMS) to simulate different pileup values.  The number of simulated events used in this analysis for each pileup value is shown in Table \ref{tab:12}.

@@

Fig. \ref{fig:tepx_cl} shows the location of TEPX clusters in Disk 4 in the XY plane (Z is the direction of the beam) from a simulated sample with a pileup of 200 \cite{dabrowski2020}. %Clusters refer to groups of hits in the pixel detector.
It consists of five concentric rings that are color-coded to indicate the density of clusters within the detector. It is transitioning from blue (low density) to red (high density), with the hottest regions indicating the overlap modules within the rings. The granularity of the data is high, with the number of bins set to 1000 for both axes and a bin size of 0.1 cm, allowing for a detailed visualization of the clusters across the disk.
%The TEPX disk in this map consists of 5 rings which are concentric rings of pixel modules at different radial distances from the beamline.

\begin{figure}[H]
  \centering
  \includegraphics[width=0.7\columnwidth]{ashish_thesis/tepx_clusters_1.png}
  \caption[TEPX Cluster Map]{TEPX cluster map in the XY plane for pileup 200 simulated sample. %Map showing coordinates of TEPX clusters in CMS global coordinate system.
    It shows clusters in one TEPX disk consisting of 5 rings. Hot regions are showing the module overlap. %for all rings (red, orange, green, yellow). Number of x bins is 1000, x bins range from -40 to 40.
    x bin size is 0.1 cm and %Number of y bins is 1000, y bins range from -40 to 40.
    y bin size is 0.1 cm.}
  \label{fig:tepx_cl}
\end{figure}

@@

\newpage
\subsection{Linearity of TEPX clusters}

Linearity is one of the systematic uncertainty in the calculation of instantaneous luminosity and PCC visible cross section $\sigma_{vis}$. A linear relation between the number of clusters and pileup (PU) imply that \\

$<N_{cl/pp}> = \frac{<N_{cl}>}{\mu} = \frac{\sigma_{vis}}{\sigma_{pp}}$ \\

$<N_{cl}> =  \frac{\sigma_{vis}}{\sigma_{pp}} \mu $ \\

Linearity indicates that the PCC visible cross section $\sigma_{vis}$ does not depend on the per bunch instantaneous luminosity (ideal luminometer). In ideal scenario, $\sigma_{vis}$ is not dependent on pileup, but this a potential problem with the luminometer. TEPX luminometer proposed for Phase II shows excellent linearity over entire pileup range. A non-linear relation $<N_{cluster}> = \alpha (PU)^{\gamma}, \gamma \neq 1$ would add non-linear terms in the rate equation $R = \sigma_{vis} L_{inst}$  and cause $\sigma_{vis}$ to vary with the per bunch instantaneous luminosity and pileup values.

The number of clusters per pp collision in Disk 4 Ring 1 varies from 0 to 2600 as a function of pileup. At pileup 200, mean number of cluster per pp collision is 2200 as shown in Fig. \ref{fig:tepx_cl_allPU}. The peak of the distribution drops as pileup increases because we have processed different number of events for different pileup values. Under low pileup conditions which is crucial for vdM calibration, number of clusters varies from 0 to 35. For average pileup value of 0.5 (vdM conditions), number of clusters varies between 0 and 12. Linearity results for simulated TEPX clusters from low to high pileup values are shown in  Fig. \ref{fig:CMS_4205} and  \ref{fig:CMS_4206}. Non-linearity is within 1 \% for entire pileup range.

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=1\columnwidth]{ashish_thesis/michigan_2.png}
  \caption[TEPX Clusters Fit And Residuals]{\onehalfspacing Left: Simulated mean number of clusters for TEPX disks as a function of pileup. A line is fitted between pileup values of 0 and 2, and then extrapolated up to a pileup of 200. Right: Deviation from linearity for clusters for  TEPX disks. The non-linearity is calculated as the relative difference between the data points and the values of the fit function at the respective pileup value. }
  \label{fig:CMS_4205}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=1\columnwidth]{ashish_thesis/michigan_3.png}
  \caption[TEPX Clusters Fit And Residuals]{\onehalfspacing Left: Simulated mean number of clusters for  TEPX Disk 4 Ring 1 as a function of pileup. A line is fitted between pileup values of 0 and 2, and then extrapolated up to a pileup of 200. Right: Deviation from linearity for clusters for TEPX Disk 4 Ring 1. The non-linearity is calculated as the relative difference between the data points and the values of the fit function at the respective pileup value. }
  \label{fig:CMS_4206}
\end{figure}

@@

Two-fold coincidences are clusters in modules overlap regions formed by modules in front and back layers of TEPX disk. There are two types of two-fold coincidences namely in $\phi$ and r. Two-fold coincidence in $\phi$ involve modules overlap in the same ring  of one double disk and two-fold coincidence in r require modules overlap between successive rings of one double disk.  Two-fold coincidences are better way to distinguish between a real hit and random electrical noise. They are more likely to be real hit than random electrical noise. Luminosity determination based on counting coincidences has an advantage over clusters that afterglow effects are tiny in the case of coincidences. 
%as shown in Fig. \ref{fig:cluster_ring_4000}. 
 %(shown in Fig. \ref{fig:cluster_ring_73})
  %as depicted in Fig. \ref{fig:cluster_ring_74}. 
 %Overlap between L1 L3 and L2 L4 layers also possible but we havent considered them in this simulation. L1L2 two fold coincidences in r are clusters in module overlaps between adjacent rings in first two layers. L3L4 two fold coincidences in r are modules overlaps in the last two layers between adjacent rings. All types of two fold coincidences in r are shown in Table \ref{tab:twofoldinr}. Module overlap between different rings for all types of two fold coincidences in r are shown in Fig. \ref{tab:twofoldinr}. Two fold coincidences are better way to distinguish between a real hit and random electrical noise. They are more likely to be real hit than random electrical noise. Luminosity determination based on counting coincidences has an advantage over clusters that afterglow effects are tiny in the case of coincidences.

@@

A new algorithm for luminosity measurement based on counting two-fold coincidences is proposed that applies selections using dr and $d\phi$ variables (shown in Fig. \ref{fig:drdphi_diagram}) to minimize non-linearity for high pileup as these variables take into account track angle and its curvature \cite{sehrawat2020}.  Proper selections are applied for each TEPX disk based on these variables using low pileup simulated samples and then these selections are extrapolated to high pileup samples for suppressing fake coincidences.

\begin{figure}[!htp]
\centering
\includegraphics[width=0.7\textwidth]{ashish_thesis/dr_dphi_variables1_1.png}
\caption[New Variables For Two Fold Coincidences]{%
  New r, phi variables used to define coincidence clusters. These variables take into account track angle and curvature.
}
\label{fig:drdphi_diagram}
\end{figure}

@@

We study the two-fold coincidences in phi for all TEPX Disk and rings using dr, dphi variable for pileup 0.5 and apply pre-selection of 0.15 cm on dr and 0.04 on dphi variable in order to remove fake coincidences.

The dr, dphi distribution showing true, fake, total two-fold coincidences for Disk 4 Ring 1 for pileup 0.5 is shown in Fig. \ref{fig:cluster_dr_dphi_dist}. After obtaining these distributions for all TEPX disks and rings, we carry out fit of the true coincidences distribution (shown in red color) to obtain mean and standard deviation of these distributions. A two-standard-deviation selection is applied on the dr value about the mean. The mean values are shown in Table \ref{tab:mdr_cuts_offset_values}. The  standard deviation values of the fit to true coincidences for all disks and rings are shown in Table \ref{tab:disk_values}. %Example fits are shown in Fig. \ref{fig:cluster_ring_1003}, \ref{fig:cluster_ring_70}, \ref{fig:cluster_ring_71}. 
Similarly, two-sigma cut is applied on the dphi variable as well. The standard deviation values for dphi variable are shown in Table \ref{tab:mdphi_cuts_values}. While doing the fit, mean value for dphi variable is set to zero. %The mean is taken as the offset value and standard deviation is used as the selection on dr, dphi variables as shown in Table \ref{tab:disk_values}, \ref{tab:mdphi_cuts_values} and \ref{tab:mdr_cuts_offset_values}. 
%The mean of the fit is taken as the offset value and standard deviation is used as the cut on dr, dphi variables.

@@

\begin{table}[h]
  \centering
  \caption[dr offset values]{dr mean values obtained from fit for two fold coincidences in phi for all disks and rings}
  \begin{tabular}{cccccc}
    \textbf{Disk} & \textbf{Ring 1} & \textbf{Ring 2} & \textbf{Ring 3} & \textbf{Ring 4} & \textbf{Ring 5} \\
    \hline
    %-1 & 0.0342041  & 0.0511416 & 0.0676943 & 0.0834355 & 0.0984932 \\
    %-2 & 0.0298276  & 0.0445398 & 0.0589828 & 0.0727454 & 0.0853587 \\
    %-3 & 0.0260326  & 0.0386608 & 0.05122   & 0.0632478 & 0.0746467 \\
    %-4 & 0.0227089  & 0.0336467 & 0.044476  & 0.0551065 & 0.06484   \\
    1  & 0.034  & 0.051 & 0.068 & 0.084 & 0.098 \\
    2  & 0.030  & 0.044 & 0.059 & 0.073 & 0.085 \\
    3  & 0.026  & 0.039 & 0.051 & 0.063 & 0.074 \\
    4  & 0.023   & 0.034 & 0.045 & 0.055 & 0.064 \\
  \end{tabular}
  \label{tab:mdr_cuts_offset_values}
\end{table}

>>>>

The Phase II upgrade of the CMS Tracker is a crucial project undertaken to maintain the effectiveness of the CMS experiment during the High Luminosity LHC (HL-LHC) phase, which is expected to start in 2029 \cite{CERN2022}. The HL-LHC will significantly increase the instantaneous luminosity, leading to a higher number of proton-proton collisions per bunch crossing. Instantaneous luminosity will increase to an unprecedented value of $7.5 \times 10^{34} cm^{-2} \: s^{-1}$ which corresponds to 200 proton-proton collisions per bunch crossing. This higher collision rate poses several challenges, including increased radiation levels, higher data rates, and more complex event environment. Run 2 pixel detector will not be able to handle these extreme radiation environment, resolve nearby particle tracks and operate properly to give a reliable estimate of the instantaneous luminosity for such a high pileup. That is why it will be replaced by a new pixel detector. The Phase II upgrade of the CMS tracker comprises of two main components: the Inner Tracker (IT) and the Outer Tracker (OT) \cite{RoyChowdhury:2729279}. The Inner Tracker consists of three main sub-components: the Tracker Endcap Pixel detector (TEPX), the Tracker Forward Pixel detector (TFPX), and the Tracker Barrel Pixel (TBPX) detector as shown in Fig. \ref{fig:Innertracker} \cite{CERN-LHCC-2017-009}.  %The Outer Tracker is responsible for detecting charged particles in the outer region of the CMS detector. It consists of two key subsystems: the silicon strip tracker and the silicon macro pixel tracker \cite{RoyChowdhury:2729279}.

Data Acquisition (DAQ) system for the Phase II CMS Tracker must handle the increased data rates and complexity associated with the HL-LHC. The HL-LHC is expected to achieve a peak luminosity which is around 5-7 times higher than the current LHC design luminosity which the DAQ system must handle. The upgraded CMS Tracker will generate a large amount of data due to the increased granularity of the pixel and strip sensors. The data rates from the tracker modules are expected to be several terabits per second (Tbps) during the HL-LHC operation. The Level-1 Trigger system in the Phase II Upgrade is designed to accept events at a rate of up to 750 kHz. The DAQ system must handle the transmission of trigger primitives to the Level-1 Trigger system and buffer the full event data until the trigger decision is made. The high-speed optical links used for data transmission between the Front-End Electronics (FEE) and off-detector electronics are expected to operate at data rates of 10-25 Gbps or higher to handle the large data volumes generated by the upgraded tracker.

The TBPX occupies the central region along the \( z \)-axis from 0 to just below 300~mm, employing 360 units of \( 1 \times 2 \) modules (modules with two readout chips) and 504 units of \( 2 \times 2 \) modules (modules with four readout chips) arranged in four layers parallel to the beam. The TFPX extends from approximately 250~mm to 1400~mm in the \( z \)-direction, with 832 units of \( 1 \times 2 \) and 896 units of \( 2 \times 2 \) modules. The TEPX covers a range from about 1700~mm to over 2600~mm and includes 768 units of \( 1 \times 2 \) and 992 units of \( 2 \times 2 \) modules. Collectively, these sub-detectors boast around 1.947 billion pixels across 13,488 pixel chips, spanning a silicon area of 4.87 square meters.

The TEPX consists of four double-disk per side with each double-disk consisting of five rings of modules as shown in Fig. \ref{fig:Innertracker_40}. %The rings have 20, 28, 36, 44 and 48 modules respectively.
The first two inner rings are equipped with \(1 \times 2\) modules, while the more extensive \(2 \times 2\) modules are mounted on the three outer rings. This design is further enhanced by the arrangement of modules in four layers as shown in Fig. \ref{fig:Innertracker_41}. This allows for modules to be mounted on both the front and back sides of a disk, creating an overlap in the r-\(\phi\) direction. By mounting modules at varying radii on the two disk that form each double-disk, overlap in the radial direction (r) is also accomplished. TEPX will have better radiation tolerance, increased granularity, improved two-track separation, improved estimation of hit rate and statistical precision, extending tracking acceptance beyond $|\eta|=4$.

@@

%TEPX will have better radiation tolerance, increased granularity, improved two-track separation, improved estimation of hit rate and statistical precision, extending tracking acceptance beyond $|\eta|=4$. Collectively, these sub-detectors boast around 1.947 billion pixels across 13,488 pixel chips, spanning a silicon area of 4.87 square meters.

@@

\begin{figure}[!htp]
  \centering
  \includegraphics[width=1 \columnwidth]{ashish_thesis/it_newgeometry_.png}
  \caption[Phase II CMS Inner Tracker]{ \onehalfspacing A layout of the CMS Phase II Inner Tracker showing four TEPX disks, eight TFPX disks and four TBPX layers \cite{Collaboration:275907420}.}
  \label{fig:Innertracker}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=0.5 \columnwidth]{ashish_thesis/tepx_DD.png}
  \caption[TEPX double disk]{ \onehalfspacing View of one TEPX double-disk in the $r-\phi$ plane; individual pixel modules on two planes (double-disk) and five rings are outlined by yellow rectangles. %The planes are separated in z by about 1 cm. The first plane holds modules on both front and back faces, with alternating rings to cover the gaps in r; the second plane holds modules in the same way as the first plane, but with an offset in $\phi$ to cover the gaps in $\phi$.
  }
  \label{fig:Innertracker_40}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=0.1 \columnwidth]{ashish_thesis/edited_tepx_alllayer_v5.png}
  \caption[TEPX double disk]{ \onehalfspacing R-Z view of the TEPX double-disk showing the four layers of modules L1, L2, L3 and L4. Layer 1 and 3 has Ring 1, 3 and 5 modules, Layer 2 and 4 has Ring 2 and Ring 4 modules. The planes are separated in z by about 1 cm. The first plane holds modules on both front and back faces, with alternating rings to cover the gaps in r; the second plane holds modules in the same way as the first plane, but with an offset in $\phi$ to cover the gaps in $\phi$.}
  \label{fig:Innertracker_41}
\end{figure}

@@

TEPX is designed to function as a high-precision online bunch-by-bunch luminometer, primarily uses pixel cluster counting for luminosity measurement, a method involving the detection and counting of pixel clusters formed by charged particles from the proton collisions. Additionally, TEPX benefits from coincidence counting due to overlapping modules which enhances accuracy and enables systematic studies. %Positioned as part of the forward extension of the Inner Tracker detector,
TEPX features a sophisticated dedicated trigger system for sampling all relevant bunch crossings in an unbiased manner. This system operates at a baseline trigger rate of 75 kHz, which is 10$\%$  of the 750 kHz allocated for physics data. %In terms of data acquisition,
TEPX adapts to different operational conditions: during standard physics runs, it maintains a baseline luminosity trigger rate of 75 kHz, while during van der Meer (vdM) scans, this rate can significantly increase, potentially up to 1 MHz \cite{Collaboration:2706512}. %TEPX readout rate (E-link speed) is estimated to be approximately 1330 Gbps at a pileup of 200  \cite{Collaboration:2706512}.
%The DAQ system of TEPX is designed to handle substantial data rates, with link bandwidth requirements varying from about 79.2 Gb/s at a trigger rate of 825 kHz during nominal physics operations to up to 24 Gb/s at the standard rate of 75 kHz  \cite{Collaboration:2706512}.

TEPX Disk 4 Ring 1 (D4R1) is located in the innermost ring of the last double-disk of the TEPX, is exclusively designed for measuring luminosity and beam-induced background (BIB). D4R1 is optimally placed due to its low radius, large active area, and high z position, making it highly efficient for these measurements. The D4R1 operational model is distinct from the rest of TEPX \cite{Haranko2023}. It employs the entire available trigger rate and readout bandwidth, which varies based on operational mode. This capability ranges from a trigger rate of 825 kHz (similar to the rest of TEPX) up to several MHz, limited by readout bandwidth and the available e-link capacity. In nominal physics conditions, the baseline trigger rate for D4R1 is 825 kHz, while in van der Meer (vdM) conditions, it is 2 MHz, with the potential to reach between 5 and 10 MHz in certain scenarios.

@@

 Simulated data samples for Phase II include full CMS detector description and uses official CMS software (version $\texttt{CMSSW\_10\_6\_0\_patch2}$) which calls GEANT4 for detector simulation \cite{GEANT4:2002zbu}.  Pixel clusters are reconstructed using the standard CMS reconstruction algorithms \cite{cmssw_2024}. It is done by converting raw data to digi format (row, column, charge) and then form clusters from digis. The samples used in simulation use single-neutrino generated events overlaid with a variable number of minimum-bias events (events with any amount of real energy detected in CMS) to simulate different pileup values.  The number of simulated events used in this analysis for each pileup value is shown in Table \ref{tab:12}.

@@

Fig. \ref{fig:tepx_cl} shows the location of TEPX clusters in Disk 4 in the XY plane (Z is the direction of the beam) from a simulated sample with a pileup of 200 \cite{dabrowski2020}. %Clusters refer to groups of hits in the pixel detector.
It consists of five concentric rings that are color-coded to indicate the density of clusters within the detector. It is transitioning from blue (low density) to red (high density), with the hottest regions indicating the overlap modules within the rings. The granularity of the histogram is set to 1000 for both axes and a bin size of 0.1 cm, allowing for a detailed visualization of the clusters across the disk.
%The TEPX disk in this map consists of 5 rings which are concentric rings of pixel modules at different radial distances from the beamline.

\begin{figure}[H]
  \centering
  \includegraphics[width=0.7\columnwidth]{ashish_thesis/tepx_clusters_1.png}
  \caption[TEPX Cluster Map]{TEPX cluster map in the XY plane for pileup 200 simulated sample having 5000 events. %Map showing coordinates of TEPX clusters in CMS global coordinate system.
    It shows clusters in one TEPX disk consisting of 5 rings. Hot regions are showing the module overlap. %for all rings (red, orange, green, yellow). Number of x bins is 1000, x bins range from -40 to 40.
    x bin size is 0.1 cm and %Number of y bins is 1000, y bins range from -40 to 40.
    y bin size is 0.1 cm.}
  \label{fig:tepx_cl}
\end{figure}

@@

\subsection{Linearity of TEPX clusters}

Linearity is one of the systematic uncertainty in the calculation of instantaneous luminosity and PCC visible cross section $\sigma_{vis}$. A linear relation between the mean number of clusters and pileup (PU) imply that $<N_{cl/pp}> = \frac{<N_{cl}>}{\mu} = \frac{\sigma_{vis}}{\sigma_{pp}}$ , $<N_{cl}> =  \frac{\sigma_{vis}}{\sigma_{pp}} \mu $. Linearity indicates that the PCC visible cross section $\sigma_{vis}$ does not depend on the per bunch instantaneous luminosity (ideal luminometer). In ideal scenario, $\sigma_{vis}$ is not dependent on pileup, but this a potential problem with the luminometer. A non-linear relation $<N_{cluster}> = \alpha (PU)^{\gamma}, \gamma \neq 1$ would add non-linear terms in the rate equation $R = \sigma_{vis} L_{inst}$ and cause $\sigma_{vis}$ to vary with the per bunch instantaneous luminosity and pileup values. TEPX luminometer proposed for Phase II shows excellent linearity over entire pileup range.

The number of clusters per pp collision in Disk 4 Ring 1 varies from 0 to 2600 as a function of pileup. At pileup 200, mean number of cluster per pp collision is 2200 as shown in Fig. \ref{fig:tepx_cl_allPU}.
%The peak of the distribution drops as pileup increases because we have processed different number of events for different pileup values. Under low pileup conditions which is crucial for vdM calibration, number of clusters varies from 0 to 35.For average pileup value of 0.5 (vdM conditions), number of clusters varies between 0 and 12.
The distributions follow Poisson statistics. Linearity results for simulated TEPX clusters from low to high pileup values are shown in  Fig. \ref{fig:CMS_4205} and  \ref{fig:CMS_4206}. Non-linearity is within 1 \% for entire pileup range.

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=1\columnwidth]{ashish_thesis/michigan_2.png}
  \caption[TEPX Clusters Fit And Residuals]{\onehalfspacing Left: Simulated mean number of clusters for TEPX disks as a function of pileup. A line is fitted between pileup values of 0 and 2, and then extrapolated up to a pileup of 200. Right: %Deviation from linearity for clusters for  TEPX disks. The non-linearity is calculated as
    The relative difference between the data points and the values of the fit function at the respective pileup value. }
  \label{fig:CMS_4205}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=1\columnwidth]{ashish_thesis/michigan_3.png}
  \caption[TEPX Clusters Fit And Residuals]{\onehalfspacing Left: Simulated mean number of clusters for  TEPX Disk 4 Ring 1 as a function of pileup. A line is fitted between pileup values of 0 and 2, and then extrapolated up to a pileup of 200. Right: %Deviation from linearity for clusters for TEPX Disk 4 Ring 1. The non-linearity is calculated as
    The relative difference between the data points and the values of the fit function at the respective pileup value. }
  \label{fig:CMS_4206}
\end{figure}

@@

Two-fold coincidences are clusters in modules overlap regions formed by modules in front layers (L1-L3) and back layers (L2-L4) separately. There are two types of two-fold coincidences namely in $\phi$ and r. Two-fold coincidence in $\phi$ involve modules overlap in the same ring of one double-disk and two-fold coincidence in r require modules overlap between successive rings of one double-disk.  Two-fold coincidences are better way to distinguish between a real hit and random noise. %They are more likely to be real hit than random electrical noise.
Luminosity determination based on counting coincidences has an advantage over clusters that afterglow effects are tiny. %in the case of coincidences. 
%as shown in Fig. \ref{fig:cluster_ring_4000}. 
 %(shown in Fig. \ref{fig:cluster_ring_73})
  %as depicted in Fig. \ref{fig:cluster_ring_74}. 
 %Overlap between L1 L3 and L2 L4 layers also possible but we havent considered them in this simulation. L1L2 two fold coincidences in r are clusters in module overlaps between adjacent rings in first two layers. L3L4 two fold coincidences in r are modules overlaps in the last two layers between adjacent rings. All types of two fold coincidences in r are shown in Table \ref{tab:twofoldinr}. Module overlap between different rings for all types of two fold coincidences in r are shown in Fig. \ref{tab:twofoldinr}. Two fold coincidences are better way to distinguish between a real hit and random electrical noise. They are more likely to be real hit than random electrical noise. Luminosity determination based on counting coincidences has an advantage over clusters that afterglow effects are tiny in the case of coincidences.

@@

A new algorithm for luminosity measurement based on counting two-fold coincidences is proposed that applies selections using dr and $d\phi$ variables (shown in Fig. \ref{fig:drdphi_diagram}) to minimize non-linearity for high pileup as these variables take into account track angle and its curvature \cite{sehrawat2020}.  Proper selections are applied for each TEPX disk based on these variables using low pileup simulated samples and then these selections are applied to high pileup samples for suppressing fake coincidences.

\begin{figure}[!htp]
\centering
\includegraphics[width=0.7\textwidth]{ashish_thesis/dr_dphi_variables1_1.png}
\caption[New Variables For Two Fold Coincidences]{%
  r, phi variables used to define coincidence clusters. %These variables take into account track angle and curvature.
}
\label{fig:drdphi_diagram}
\end{figure}

@@

We study the two-fold coincidences in $\phi$ for all TEPX Disk and rings using r, phi variable for pileup 0.5. Based on pileup 0.5 sample, a pre-selection of 0.15 cm on dr and 0.04 on dphi variable is applied in order to remove fake coincidences.

The dr, dphi distribution showing true, fake, total two-fold coincidences for Disk 4 Ring 1 for pileup 0.5 is shown in Fig. \ref{fig:cluster_dr_dphi_dist}. After obtaining these distributions for all TEPX disks and rings, we carry out fit of the true coincidences distribution (shown in red color) to obtain mean and standard deviation of these distributions. A two-standard-deviation selection is applied on the dr value about the mean. The mean values are shown in Table \ref{tab:mdr_cuts_offset_values}. The  standard deviation values of the fit to true coincidences for all disks and rings are shown in Table \ref{tab:disk_values}. %Example fits are shown in Fig. \ref{fig:cluster_ring_1003}, \ref{fig:cluster_ring_70}, \ref{fig:cluster_ring_71}. 
Similarly, two-sigma cut is applied on the dphi variable as well. The standard deviation values for dphi variable are shown in Table \ref{tab:mdphi_cuts_values}. While doing the fit, mean value for dphi variable is set to zero. %The mean is taken as the offset value and standard deviation is used as the selection on dr, dphi variables as shown in Table \ref{tab:disk_values}, \ref{tab:mdphi_cuts_values} and \ref{tab:mdr_cuts_offset_values}. 
%The mean of the fit is taken as the offset value and standard deviation is used as the cut on dr, dphi variables.
A plot of two-fold coincidences in phi after applying all selection based on standard deviation values of all TEPX disks and rings is shown in Fig. \ref{fig:cluster_ring_720}.

@@

\begin{table}[H]
  \centering
  \caption[dr offset values]{dr mean values obtained from fit for two fold coincidences in phi for all disks and rings}
  \begin{tabular}{cccccc}
    \textbf{Disk} & \textbf{Ring 1} & \textbf{Ring 2} & \textbf{Ring 3} & \textbf{Ring 4} & \textbf{Ring 5} \\
    \hline
    %-1 & 0.0342041  & 0.0511416 & 0.0676943 & 0.0834355 & 0.0984932 \\
    %-2 & 0.0298276  & 0.0445398 & 0.0589828 & 0.0727454 & 0.0853587 \\
    %-3 & 0.0260326  & 0.0386608 & 0.05122   & 0.0632478 & 0.0746467 \\
    %-4 & 0.0227089  & 0.0336467 & 0.044476  & 0.0551065 & 0.06484   \\
    1  & 0.034  & 0.051 & 0.068 & 0.084 & 0.098 \\
    2  & 0.030  & 0.044 & 0.059 & 0.073 & 0.085 \\
    3  & 0.026  & 0.039 & 0.051 & 0.063 & 0.074 \\
    4  & 0.023   & 0.034 & 0.045 & 0.055 & 0.064 \\
  \end{tabular}
  \label{tab:mdr_cuts_offset_values}
\end{table}
